Reanalysis Step 3: Propensity Score Analysis with Twang

### 2025-08-19 reanalysis-3-propensity_scores.qmd
cleanup 3

```{r setup}
#| include: false
# Setup analysis environment
source("dependencies.R")
setup_analysis(seed = 2025)

@@

```{r load-data}
# Load the imputed datasets
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_imputations <- length(imputed_datasets)

@@

# Load variable lists
variable_lists <- readRDS(file.path(reanalysis_data_dir, "variable_lists.rds"))
```

@@

  # Simple approach: use summary to assess balance
  # The summary() function prints balance info but may not return it in a usable format
  # So we'll calculate basic balance metrics manually

@@

  # Calculate weighted balance for a few key variables
  # This is a simplified assessment
  balance_summaries[[i]] <- data.frame(
    Imputation = i,
    N_Treated = sum(treated),
    N_Control = sum(!treated),
    Mean_Weight_Treated = mean(weights[treated]),
    Mean_Weight_Control = mean(weights[!treated]),
    Max_Weight = max(weights)
  )

  # Print summary for diagnostic purposes (this will show balance)
  if (i <= max_imputations_for_diagnostics) {
    message("Balance summary for imputation ", i, ":")
    print(summary(twang_model))
  }
}

@@

# Check weight stability
if (avg_weights$Max_Weight_Overall > 10) {
  message("⚠ Warning: Maximum weight > 10, indicating potential instability")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability")
} else {
  message("✓ Weights appear stable (max < 5)")
}

@@

# Select imputations to show diagnostics for
n_to_show <- min(max_imputations_for_diagnostics, n_imputations)
imps_to_show <- 1:n_to_show

@@

  # Get the twang model for this imputation
  twang_model <- twang_results[[i]]$model

  # Try to generate diagnostic plots
  tryCatch({
    # The standard twang plots
    plot(twang_model)
  }, error = function(e) {
    message("  Standard plots not available, generating custom diagnostics...")

    # Create custom diagnostic plots
    ps_scores <- twang_results[[i]]$ps_scores
    weights <- twang_results[[i]]$weights
    treated <- imputed_datasets[[i]]$ever_lapse_binary == 1

@@

    # Plot 1: PS distribution by treatment
    hist(ps_scores[!treated], main = "PS Distribution - Control", 
         xlab = "Propensity Score", col = "lightblue")
    hist(ps_scores[treated], main = "PS Distribution - Treated", 
         xlab = "Propensity Score", col = "lightcoral")

    # Plot 2: Weight distribution
    hist(weights[!treated], main = "Weights - Control", 
         xlab = "Weight", col = "lightblue")
    hist(weights[treated], main = "Weights - Treated", 
         xlab = "Weight", col = "lightcoral")

@@

  # Print summary
  print(summary(twang_model))
}
```

@@

  # Get propensity scores from the twang model
  twang_model <- twang_results[[i]]$model
  weights <- twang_results[[i]]$weights
  treated <- imputed_datasets[[i]]$ever_lapse_binary == 1

  # For twang models, propensity scores may not be directly accessible
  # We can use the weights to show distribution instead
  # Since weights = 1/PS for treated and 1/(1-PS) for control in ATT weighting

  # Create histograms of weights instead of propensity scores
  par(mfrow = c(1, 2))

@@

```{r save-results}
# Save twang PS results for use in outcome analysis
ps_output <- list(
  twang_results = twang_results,
  weight_summary = balance_summary_all,
  avg_weights = avg_weights,
  formula_used = ps_formula,
  n_imputations = n_imputations,
  method = "twang_gbm",
  n_trees = n_trees,
  analysis_mode = analysis_mode
)

@@

```{r summary}
# Final summary
message("=== TWANG PROPENSITY SCORE ANALYSIS SUMMARY ===")
message("Method: Generalized Boosted Models (GBM) via twang")
message("Imputations processed: ", n_imputations)
message("Analysis mode: ", analysis_mode)
message("Trees used: ", n_trees)
message("Estimand: ATT (Average Treatment Effect on the Treated)")
message("Stop method: es.mean (effect size minimization)")

@@

# Weight stability assessment
if (avg_weights$Max_Weight_Overall > 10) {
  message("⚠ Warning: Maximum weight > 10, indicating potential instability")
  message("  Consider increasing n.trees or adjusting twang parameters")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability in outcome analysis")
} else {
  message("✓ Weights appear stable (max < 5)")
}

>>>>

```{r setup}
#| include: false
source("dependencies.R")
setup_analysis(seed = 2025)

@@

```{r load-data}
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_imputations <- length(imputed_datasets)

@@

variable_lists <- readRDS(file.path(reanalysis_data_dir, "variable_lists.rds"))
```

@@

@@

  balance_summaries[[i]] <- data.frame(
    Imputation = i,
    N_Treated = sum(treated),
    N_Control = sum(!treated),
    Mean_Weight_Treated = mean(weights[treated]),
    Mean_Weight_Control = mean(weights[!treated]),
    Max_Weight = max(weights)
  )

  if (i <= max_imputations_for_diagnostics) {
    message("Balance summary for imputation ", i, ":")
    print(summary(twang_model))
  }
}

@@

if (avg_weights$Max_Weight_Overall > 10) {
  message("Warning: Maximum weight > 10, indicating potential instability")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability")
} else {
  message("Weights appear stable (max < 5)")
}

@@

n_to_show <- min(max_imputations_for_diagnostics, n_imputations)
imps_to_show <- 1:n_to_show

@@

  twang_model <- twang_results[[i]]$model

  tryCatch({
    plot(twang_model)
  }, error = function(e) {
    message("  Standard plots not available, generating custom diagnostics...")

    ps_scores <- twang_results[[i]]$ps_scores
    weights <- twang_results[[i]]$weights
    treated <- imputed_datasets[[i]]$ever_lapse_binary == 1

@@

    hist(ps_scores[!treated], main = "PS Distribution - Control", 
         xlab = "Propensity Score", col = "lightblue")
    hist(ps_scores[treated], main = "PS Distribution - Treated", 
         xlab = "Propensity Score", col = "lightcoral")

    hist(weights[!treated], main = "Weights - Control", 
         xlab = "Weight", col = "lightblue")
    hist(weights[treated], main = "Weights - Treated", 
         xlab = "Weight", col = "lightcoral")

@@

  print(summary(twang_model))
}
```

@@

  twang_model <- twang_results[[i]]$model
  weights <- twang_results[[i]]$weights
  treated <- imputed_datasets[[i]]$ever_lapse_binary == 1

  par(mfrow = c(1, 2))

@@

```{r save-results}
ps_output <- list(
  twang_results = twang_results,
  weight_summary = balance_summary_all,
  avg_weights = avg_weights,
  formula_used = ps_formula,
  n_imputations = n_imputations,
  method = "twang_gbm",
  n_trees = n_trees,
  analysis_mode = analysis_mode
)

@@

```{r summary}
message("Analysis Summary:")
message("Method: Generalized Boosted Models (GBM) via twang")
message("Imputations processed: ", n_imputations)
message("Analysis mode: ", analysis_mode)
message("Trees used: ", n_trees)
message("Estimand: ATT (Average Treatment Effect on the Treated)")
message("Stop method: es.mean (effect size minimization)")

@@

# Weight stability assessment
if (avg_weights$Max_Weight_Overall > 10) {
  message("  Warning: Maximum weight > 10, indicating potential instability")
  message("  Consider increasing n.trees or adjusting twang parameters")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability in outcome analysis")
} else {
  message("Weights appear stable (max < 5)")
}
